Suitability prevalence area index in the late Quaternary explains genetic diversity in Tassel-eared Squirrels (\textit{Sciurus aberti})

### 2023-08-01 manuscript_english.tex
Additional edits to Intro and start of framework

\title{Suitability prevalence area index in late quaternary explains
genetic diversity in Tassel eared Squirrels}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

In this work, we propose a method to find what we call the Suitability Prevalence Area
(SPA) that is the time-averaged area of highest environmental suitability for a given species. The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of the target species; and 2) predicting the expected patterns of genetic diversity given the inferred history of population stability and connectivity.

To obtain the SPA, we overlapped historical reconstructions of the
geographical distribution of a target species estimated every 2,000 years from the present day until 120,000 years ago, in each time interval recording the environmental suitability at each modeled site.

Subsequently, we delimited historical endemic areas to locations where
the prevalence of suitability remained at 90\% during this period. From
the fixation index calculated with respect to populations in the
historical endemic areas, a statistical model of the fixation index as a
function of SPA was performed. With this statistical model, the fixation
index values were projected to the current distribution to obtain a map
with the geographic pattern of this index.

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates evolutionary fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of evolutionary fitness that depends on
environmental conditions.

That is, the sites in the geography \(\mathbf{G}_o\) that present
positive growth rates are related to favorable conditions. If we take
the set of places in ambient space where these values are greater than a
threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, that
is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

This is a way of expressing the fundamental niche as a function of
environmental conditions. So we can map each point of geography to
ambient space. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive rates. However, we assume that the environment
is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environment and
therefore is different at each time.

Thus locally the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic rate
that depends on the change in environmental conditions over time is:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

@@

As a case study we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
(Bono et al. 2018; Burgin et al. 2018). In order to select this species,
we list the following considerations:

>>>>

\title{Suitability prevalence area index in the late Quaternary explains genetic diversity in Tassel-eared Squirrels (\textit{Sciurus aberti})}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

In this work, we propose a method to find what we call the Suitability Prevalence Area
(SPA) that is the time-averaged area of highest environmental suitability for a given species, defined as locations with ≥ 90\% suitability over a historical time period. The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) predicting the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.

As a case study, we consider Tassel-eared Squirrels (\emph{Sciurus aberti}), a species of rodent that is currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution (Bono et al. 2018; Burgin et al. 2018). To obtain the SPA for \textit{S. aberti}, we overlapped historical reconstructions of the species' geographical distribution estimated every 2,000 years from the present day until 120,000 years ago, recording the environmental suitability in each time interval at each modeled site.

Subsequently, we delimited historical endemic areas to locations where
the prevalence of suitability remained at 90\% during this period.

Subsequently, we calculated the fixation index (\(F_{st}\)) of the genetic relatedness among populations and compared that to the SPA index of historical endemism, constructing a statistical model of the fixation index as a function of the SPA. With this statistical model, the \(F_{st}\) values were projected to the current geographic distribution of \textit{S. aberti} to obtain a predictive map of expected \(F_{st}\) values over space.

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates evolutionary fitness to environment. That is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of evolutionary fitness that depends on
environmental conditions.

In other words, the sites in the geography \(\mathbf{G}_o\) that present
positive population growth rates are related to favorable environmental conditions. If we take
the set of places in ambient space where these values are greater than a
certain threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, as \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

This is a way of expressing the fundamental niche \(\mathbf{N}_F\) as a function of
environmental conditions. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However, all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive growth rates. Nonetheless, we assume that the environment
is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environmental conditions and
therefore growth rate changes in value at each time interval.

Locally, the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

@@

As a case study, we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
(Bono et al. 2018; Burgin et al. 2018). In order to select this species,
we list the following considerations:
